---
title: "IMRS Step 09 Evaluation Report"
author: "Yansheng Luo"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 70),
  warning = FALSE,
  message = FALSE
)

library(readr)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)

# ------------------------------
# Global plotting style
# ------------------------------

# Professional color palette
imrs_palette <- c(
  anchor = "#4C78A8",
  external = "#F58518"
)

condition_palette <- c(
  CONTROL = "#9AA0A6",
  DELIVERY = "#D64D4D"
)

# Global ggplot theme
theme_set(
  theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(
        size = 14,
        face = "bold",
        hjust = 0
      ),
      plot.subtitle = element_text(
        size = 11
      ),
      axis.title = element_text(
        size = 12,
        face = "bold"
      ),
      axis.text = element_text(
        size = 10
      ),
      legend.title = element_text(
        size = 11,
        face = "bold"
      ),
      legend.text = element_text(
        size = 10
      ),
      legend.position = "right",
      panel.border = element_blank(),
      axis.line = element_line(linewidth = 0.6),
      plot.margin = margin(10, 12, 10, 10)
    )
)

# Make reference lines look cleaner
update_geom_defaults(
  "vline",
  list(color = "grey40", linewidth = 0.7)
)

update_geom_defaults(
  "hline",
  list(color = "grey40", linewidth = 0.7)
)

update_geom_defaults(
  "point",
  list(size = 2.4)
)

update_geom_defaults(
  "line",
  list(linewidth = 1)
)

update_geom_defaults(
  "errorbarh",
  list(linewidth = 0.7)
)

update_geom_defaults(
  "col",
  list(fill = "#4C78A8")
)
```

# 1.\ Overview

This report summarizes the biological and cross-dataset performance of the Innate Immune Response Score (IMRS) using the Step 09 split-contrast evaluation outputs.

Main questions:

1. Does IMRS separate DELIVERY from CONTROL samples?
2. Does the score generalize beyond anchor datasets?
3. Are the frozen score weights biologically plausible?

The analysis below uses Step 09 evaluation tables and the frozen anchor gene weights.

```{r load-data}
project_root <- "D:/IMRS_Project"

eval_tbl <- read_tsv(file.path(project_root, "05_score/transfer/eval/step09_split_eval.tsv"))
summary_tbl <- read_tsv(file.path(project_root, "05_score/transfer/eval/step09_split_summary.tsv"))
sample_tbl <- read_tsv(file.path(project_root, "05_score/transfer/eval/step09_split_sample_level.tsv"))
weights_tbl <- read_tsv(file.path(project_root, "05_score/anchors/gene_weights.tsv"))
```

```{r checks}
required_eval_cols <- c(
  "pass","dataset_type","gse_id","split_id",
  "delta_mean_imrs_z","auc_imrs_z"
)

missing_eval <- setdiff(required_eval_cols, names(eval_tbl))

if(length(missing_eval) > 0){
  stop("eval_tbl missing columns: ", paste(missing_eval, collapse=", "))
}

if(!"imrs_z" %in% names(sample_tbl)){
  stop("sample_tbl must contain imrs_z")
}

if(!"condition_simple" %in% names(sample_tbl)){
  stop("sample_tbl must contain condition_simple")
}
```

```{r derive-passing-data}
pass_eval <- eval_tbl %>%
  dplyr::filter(pass)

join_keys <- intersect(c("gse_id","split_id","dataset_type"), names(sample_tbl))

if(length(join_keys) >= 2){
  pass_sample_tbl <- sample_tbl %>%
    dplyr::inner_join(
      pass_eval %>% dplyr::select(all_of(join_keys)) %>% dplyr::distinct(),
      by = join_keys
    )
} else {
  pass_sample_tbl <- sample_tbl
}
```

# 2.\ Basic summary statistics

We summarize overall Step 09 performance across passing contrasts.

```{r overall-summary}
overall_stats <- pass_eval %>%
  summarise(
    n_contrasts = n(),
    mean_delta = mean(delta_mean_imrs_z, na.rm=TRUE),
    median_delta = median(delta_mean_imrs_z, na.rm=TRUE),
    prop_positive_delta = mean(delta_mean_imrs_z > 0, na.rm=TRUE),
    mean_auc = mean(auc_imrs_z, na.rm=TRUE),
    median_auc = median(auc_imrs_z, na.rm=TRUE)
  )

overall_stats
```

Positive `delta_mean_imrs_z` indicates DELIVERY samples have higher IMRS scores than CONTROL samples.  
AUC values above 0.5 indicate separation ability.

# 3.\ Distribution of effect size across contrasts

Primary metric:

\[
\Delta IMRS_z =
\overline{IMRS_z(DELIVERY)} -
\overline{IMRS_z(CONTROL)}
\]

```{r delta-distribution, fig.width=8, fig.height=5}

ggplot(
  pass_eval,
  aes(x = delta_mean_imrs_z, fill = dataset_type)
) +
  geom_histogram(bins = 30, alpha = 0.8, color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ dataset_type, scales = "free_y") +
  scale_fill_manual(values = imrs_palette) +
  labs(
    title = "Distribution of IMRS activation effect size",
    subtitle = "Anchor and external datasets shown separately",
    x = expression(Delta*" mean IMRS"[z]*" (DELIVERY - CONTROL)"),
    y = "Number of contrasts"
  )
```

Anchor datasets contributed to construction of the IMRS score and therefore represent in-sample behavior. External datasets provide a more stringent test of score transferability. Separating the two groups avoids overstating generalization performance and makes differences between training and validation datasets easier to interpret.

# 3B. Contrast-level effects by dataset

This plot shows each DELIVERY–CONTROL contrast individually.  
Large vertical spread within a dataset indicates heterogeneous biological responses.

```{r contrast-scatter, fig.width=8, fig.height=6}
ggplot(
  pass_eval,
  aes(
    x = delta_mean_imrs_z,
    y = reorder(gse_id, delta_mean_imrs_z, FUN = median),
    color = dataset_type
  )
) +
  geom_point(alpha = 0.7, size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(
    title = "Contrast-level IMRS activation across datasets",
    x = expression(Delta*" mean IMRS"[z]*" (DELIVERY - CONTROL)"),
    y = "Dataset (GSE ID)",
    color = "Dataset type"
  )
```

# 3B. Contrast-level effects by dataset

This plot shows each DELIVERY–CONTROL contrast individually.  
Large vertical spread within a dataset indicates heterogeneous biological responses.

```{r contrast-scatter, fig.width=8, fig.height=6}
ggplot(
  pass_eval,
  aes(
    x = delta_mean_imrs_z,
    y = reorder(gse_id, delta_mean_imrs_z, FUN = median),
    color = dataset_type
  )
) +
  geom_point(alpha = 0.7, size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(
    title = "Contrast-level IMRS activation across datasets",
    x = expression(Delta*" mean IMRS"[z]*" (DELIVERY - CONTROL)"),
    y = "Dataset (GSE ID)",
    color = "Dataset type"
  )
```

# 4.\ Anchor versus external datasets

```{r anchor-summary}
anchor_external_stats <- pass_eval %>%
  group_by(dataset_type) %>%
  summarise(
    n_contrasts = n(),
    mean_delta = mean(delta_mean_imrs_z, na.rm=TRUE),
    median_delta = median(delta_mean_imrs_z, na.rm=TRUE),
    prop_positive_delta = mean(delta_mean_imrs_z > 0, na.rm=TRUE),
    mean_auc = mean(auc_imrs_z, na.rm=TRUE),
    median_auc = median(auc_imrs_z, na.rm=TRUE),
    .groups="drop"
  )

anchor_external_stats
```

```{r anchor-vs-external, fig.width=7, fig.height=5}
ggplot(
  pass_eval,
  aes(x = dataset_type, y = delta_mean_imrs_z, fill = dataset_type)
) +
  geom_boxplot(alpha=0.8) +
  geom_hline(yintercept = 0, linetype="dashed") +
  theme_classic() +
  labs(
    title="IMRS activation in anchor versus external datasets",
    x="Dataset type",
    y=expression(Delta*" mean IMRS"[z])
  )
```

Anchors are expected to show stronger effects because they contributed to score construction.

# 5.\ Classification performance across contrasts

```{r auc-distribution, fig.width=8, fig.height=5}
ggplot(
  pass_eval %>% filter(is.finite(auc_imrs_z)),
  aes(x = auc_imrs_z, fill = dataset_type)
) +
  geom_histogram(bins=30, alpha=0.7, position="dodge") +
  geom_vline(xintercept=0.5, linetype="dashed") +
  theme_classic() +
  labs(
    title="AUC distribution across contrasts",
    x="AUC",
    y="Number of contrasts",
    fill="Dataset type"
  )
```

High AUC indicates strong classification ability. However, small sample sizes may inflate AUC values.

# 6.\ dataset-level score separation

```{r dataset-level-separation, fig.width=8, fig.height=6}


dataset_sep <- pass_eval %>%
  dplyr::group_by(gse_id, dataset_type) %>%
  dplyr::summarise(
    mean_delta = mean(delta_mean_imrs_z, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  dataset_sep,
  aes(x = mean_delta, y = reorder(gse_id, mean_delta), fill = dataset_type)
) +
  geom_col(width = 0.7, color = "black", linewidth = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = imrs_palette) +
  theme_classic() +
  labs(
    title = "Dataset-level IMRS separation",
    subtitle = "Average DELIVERY minus CONTROL score difference per dataset",
    x = expression("Mean " * Delta * " IMRS"[z]),
    y = "Dataset",
    fill = "Dataset type"
  )
```

Because IMRS values are standardized within each dataset using dataset-specific control samples, absolute sample-level values are not directly comparable across studies. For this reason, pooled sample-level boxplots were replaced with a dataset-level summary of DELIVERY–CONTROL separation. Positive values indicate higher IMRS activation in DELIVERY samples within the same dataset.

# 7.\ Effect size versus classification strength

```{r delta-vs-auc, fig.width=7, fig.height=5}
scatter_df <- pass_eval %>%
  filter(is.finite(delta_mean_imrs_z), is.finite(auc_imrs_z)) %>%
  mutate(sample_size = n_controls + n_delivery)

ggplot(
  scatter_df,
  aes(
    x = delta_mean_imrs_z,
    y = auc_imrs_z,
    color = dataset_type,
    size = sqrt(sample_size)
  )
) +
  geom_point(alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  facet_wrap(~ dataset_type) +
  scale_color_manual(values = imrs_palette) +
  labs(
    title = "Effect size vs classification accuracy",
    subtitle = "Anchor and external datasets shown separately",
    x = expression(Delta*" mean IMRS"[z]),
    y = "AUC",
    size = "sqrt(sample size)"
  )
```

Both effect size ($\Delta \mathrm{IMRS}_z$) and AUC measure separation between DELIVERY and CONTROL samples. Faceting by dataset type allows direct comparison between anchor datasets used during score construction and independent external datasets used for validation. Point size reflects total sample size, highlighting contrasts supported by larger datasets.

# 8.\ Top contrasts by performance

```{r top-contrasts}
# Select the single best-performing contrast from each dataset
top_contrasts <- pass_eval %>%
  group_by(gse_id) %>%
  slice_max(
    order_by = delta_mean_imrs_z,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  arrange(desc(delta_mean_imrs_z)) %>%
  select(
    dataset_type,
    gse_id,
    split_id,
    tissue,
    time_h,
    n_controls,
    n_delivery,
    delta_mean_imrs_z,
    auc_imrs_z
  ) %>%
  slice_head(n = 15)

top_contrasts
```

# 9.\ Top weighted genes in IMRS

```{r top-genes-prepare}
# install/load biomaRt only if needed
if (!requireNamespace("biomaRt", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  BiocManager::install("biomaRt", ask = FALSE, update = FALSE)
}

library(biomaRt)

mapping_file <- file.path(project_root, "05_score/anchors/gene_symbol_mapping.tsv")

# determine gene column
if ("gene_symbol" %in% names(weights_tbl)) {
  gene_col <- "gene_symbol"
  note <- "Gene symbols"
} else if ("gene_id" %in% names(weights_tbl)) {
  gene_col <- "gene_id"
  note <- "Mapped from Ensembl IDs"
} else if ("gene" %in% names(weights_tbl)) {
  gene_col <- "gene"
  note <- "Generic gene column"
} else {
  stop("weights file missing gene column")
}

if (!"beta_meta" %in% names(weights_tbl)) {
  stop("weights file missing beta_meta")
}

top_genes <- weights_tbl %>%
  dplyr::mutate(
    gene_id_raw = as.character(.data[[gene_col]]),
    gene_id_clean = sub("\\..*$", "", gene_id_raw)
  ) %>%
  dplyr::arrange(desc(abs(beta_meta))) %>%
  dplyr::slice_head(n = 25)

download_mouse_mapping <- function(gene_ids) {
  mart <- useEnsembl(
    biomart = "genes",
    dataset = "mmusculus_gene_ensembl"
  )

  mapping <- getBM(
    attributes = c("ensembl_gene_id", "mgi_symbol"),
    filters = "ensembl_gene_id",
    values = unique(gene_ids),
    mart = mart
  )

  mapping %>%
    dplyr::mutate(
      ensembl_gene_id = as.character(ensembl_gene_id),
      mgi_symbol = as.character(mgi_symbol)
    )
}

is_ensembl <- any(grepl("^ENS", top_genes$gene_id_clean))

if (is_ensembl) {

  need_redownload <- TRUE

  if (file.exists(mapping_file)) {
    mapping_try <- tryCatch(
      read_tsv(mapping_file, show_col_types = FALSE),
      error = function(e) NULL
    )

    if (!is.null(mapping_try) &&
        all(c("ensembl_gene_id", "mgi_symbol") %in% names(mapping_try))) {

      mapping_try <- mapping_try %>%
        dplyr::mutate(
          ensembl_gene_id = as.character(ensembl_gene_id),
          mgi_symbol = as.character(mgi_symbol)
        )

      if (nrow(mapping_try) > 0 &&
          any(!is.na(mapping_try$ensembl_gene_id)) &&
          any(mapping_try$ensembl_gene_id != "")) {
        mapping <- mapping_try
        need_redownload <- FALSE
      }
    }
  }

  if (need_redownload) {
    all_gene_ids_clean <- weights_tbl[[gene_col]] %>%
      as.character() %>%
      sub("\\..*$", "", .)

    mapping <- download_mouse_mapping(all_gene_ids_clean)
    dir.create(dirname(mapping_file), recursive = TRUE, showWarnings = FALSE)
    write_tsv(mapping, mapping_file)
  }

  top_genes <- top_genes %>%
    dplyr::left_join(
      mapping,
      by = c("gene_id_clean" = "ensembl_gene_id")
    ) %>%
    dplyr::mutate(
      gene_label = ifelse(
        is.na(mgi_symbol) | mgi_symbol == "",
        gene_id_clean,
        mgi_symbol
      )
    )

} else {

  top_genes <- top_genes %>%
    dplyr::mutate(gene_label = gene_id_raw)
}

top_genes %>%
  dplyr::select(gene_id_raw, gene_id_clean, gene_label, beta_meta)
```

```{r top-genes-plot, fig.width=8, fig.height=7}
ggplot(
  top_genes,
  aes(x = reorder(gene_label, beta_meta), y = beta_meta)
) +
  geom_col() +
  coord_flip() +
  theme_classic() +
  labs(
    title = "Top weighted IMRS genes",
    subtitle = note,
    x = "Gene",
    y = "Meta-weight"
  )
```

The highest-weighted IMRS genes include canonical interferon-stimulated and inflammatory response genes such as *Ifit1, Ifit3, Mx2, Cxcl10, Cxcl11, Ccl2, Ccl7, Gbp10, Gbp11,* and *Acod1*. This strongly supports the biological interpretation that IMRS captures delivery-associated innate immune activation rather than an arbitrary technical signal.

# 10.\ Mean performance by dataset

```{r dataset-summary}
dataset_summary <- summary_tbl %>%
  arrange(desc(mean_delta_mean_imrs_z))

dataset_summary
```

Datasets with weaker or negative signal should be examined individually.

# 11.\ Forest-style summary by dataset

This plot summarizes average effect size by dataset with approximate 95% confidence intervals across passing contrasts.

```{r forest-dataset-plot, fig.width=8, fig.height=6}
dataset_forest <- pass_eval %>%
  group_by(gse_id, dataset_type) %>%
  summarise(
    n_contrasts = n(),
    mean_delta = mean(delta_mean_imrs_z, na.rm = TRUE),
    sd_delta = sd(delta_mean_imrs_z, na.rm = TRUE),
    se_delta = sd_delta / sqrt(n_contrasts),
    lower = mean_delta - 1.96 * se_delta,
    upper = mean_delta + 1.96 * se_delta,
    .groups = "drop"
  ) %>%
  arrange(mean_delta)

dataset_forest$gse_id <- factor(
  dataset_forest$gse_id,
  levels = unique(dataset_forest$gse_id)
)

ggplot(
  dataset_forest,
  aes(x = mean_delta, y = gse_id, color = dataset_type)
) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_point(size = 2.5) +
  theme_classic() +
  labs(
    title = "Mean IMRS effect size by dataset",
    x = expression("Mean " * Delta * " IMRS"[z] * " with 95% CI"),
    y = "Dataset",
    color = "Dataset type"
  )
```

This plot makes it easier to see which datasets consistently support the score and which datasets are closer to zero.

# 12.\ Distribution of gene weights

This plot shows the global distribution of frozen IMRS weights.

```{r gene-weight-distribution, fig.width=8, fig.height=5}
ggplot(
  weights_tbl,
  aes(x = beta_meta)
) +
  geom_histogram(bins = 40, alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(
    title = "Distribution of frozen IMRS gene weights",
    x = "Meta-weight",
    y = "Number of genes"
  )
```

A roughly centered distribution with tails in both directions is expected for a weighted score, while very extreme skew could indicate instability or over-dominance by a small number of genes.

# 13.\ Example ROC curve

If sample-level truth labels are available, we can plot one example ROC curve for the best-performing passing contrast.

```{r roc-example-prepare}
roc_ready <- all(c("gse_id","split_id","condition_simple","imrs_z") %in% names(pass_sample_tbl))

if(roc_ready){
  best_row <- pass_eval %>%
    filter(is.finite(auc_imrs_z)) %>%
    arrange(desc(auc_imrs_z), desc(delta_mean_imrs_z)) %>%
    slice(1)

  roc_df <- pass_sample_tbl %>%
    filter(
      gse_id == best_row$gse_id,
      split_id == best_row$split_id,
      condition_simple %in% c("CONTROL","DELIVERY")
    ) %>%
    mutate(
      truth = ifelse(condition_simple == "DELIVERY", 1, 0)
    ) %>%
    filter(is.finite(imrs_z))
}
```

```{r roc-example-plot, fig.width=6, fig.height=6}
if(roc_ready && exists("roc_df") && nrow(roc_df) > 2 && length(unique(roc_df$truth)) == 2){

  thresholds <- sort(unique(roc_df$imrs_z), decreasing = TRUE)

  roc_curve <- lapply(thresholds, function(thr){
    pred <- ifelse(roc_df$imrs_z >= thr, 1, 0)
    tp <- sum(pred == 1 & roc_df$truth == 1)
    fp <- sum(pred == 1 & roc_df$truth == 0)
    tn <- sum(pred == 0 & roc_df$truth == 0)
    fn <- sum(pred == 0 & roc_df$truth == 1)

    tibble(
      threshold = thr,
      tpr = ifelse((tp + fn) > 0, tp / (tp + fn), NA_real_),
      fpr = ifelse((fp + tn) > 0, fp / (fp + tn), NA_real_)
    )
  }) %>%
    bind_rows() %>%
    distinct(fpr, tpr, .keep_all = TRUE) %>%
    arrange(fpr, tpr)

  roc_curve <- bind_rows(
    tibble(threshold = Inf, fpr = 0, tpr = 0),
    roc_curve,
    tibble(threshold = -Inf, fpr = 1, tpr = 1)
  ) %>%
    distinct(fpr, tpr, .keep_all = TRUE) %>%
    arrange(fpr, tpr)

  ggplot(roc_curve, aes(x = fpr, y = tpr)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    geom_line(linewidth = 1) +
    geom_point(size = 1.5) +
    coord_equal() +
    theme_classic() +
    labs(
      title = paste0("Example ROC curve: ", best_row$gse_id, " / ", best_row$split_id),
      x = "False positive rate",
      y = "True positive rate"
    )

} else {
  plot.new()
  text(0.5, 0.5, "ROC example unavailable:\nrequired sample-level columns or class balance missing.")
}
```

This example provides a visual demonstration of classification behavior for one of the strongest passing contrasts.

# 14.\ Caveats

Key limitations:

1. Some contrasts have small sample sizes.
2. AUC can be unstable with small N.
3. External datasets may vary in delivery method and tissue context.
4. Gene identifiers in the weights table require mapping to symbols for biological interpretation.

# 15.\ Conclusion

The Step 09 evaluation suggests that IMRS captures delivery-associated innate immune activation across multiple studies.

Supporting evidence includes:

1. Most contrasts show positive effect size.
2. Many contrasts show AUC above 0.5.
3. DELIVERY samples tend to have higher IMRS scores than CONTROL samples.
4. External datasets retain detectable signal.

Overall, IMRS appears technically functional and shows promising transferability across datasets, although further biological validation is required.
```
